---
title: "Egg offspring count data analysis"
author: "Kiran Adhikari"
date: "7/3/2025"
output:
   github_document
---


#### Load required packages
```{r load-libs, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyverse)
library(ggpubr)
library(lme4)
library(emmeans)
```


#### Read data and calculate the eggs and offspring per female per tube per day
```{r}
setwd("~/Documents/Data/Ovariole_number/")
Data<- read.csv("Egg_offspring_count_data.csv", header=TRUE)
Data<- Data %>% mutate(across(c('Block','Day', 'Genotype', 'Treatment', 'Tube'), as.factor))
colnames(Data)[6]<- "Number_of_females"
colnames(Data)[7]<- "Number_of_eggs"
colnames(Data)[8]<- "Number_of_offspring"
Data$Eggs_per_female<- Data$Number_of_eggs/Data$Number_of_females
Data$Offspring_per_female<- Data$Number_of_offspring/Data$Number_of_females
```


#### Check if any columns have NA values
```{r}
apply(Data, 2, function(x) any(is.nan(x)))
```


#### Calculate average number of eggs and offsprings per female for each tube across all 5 days
```{r}
Data_updated<- Data %>% group_by(Genotype, Treatment, Tube, Block) %>% 
        mutate(Total_eggs = sum(Eggs_per_female), Total_offspring = sum(Offspring_per_female)) %>% distinct(Genotype, Treatment, Tube, Total_eggs, Total_offspring)
```


### Egg data

#### First arrange the genotypes according to their ovariole number and plot the egg numbers
```{r}
library(ggplot2)
level_order<- c('RAL 837', 'RAL 397', 'RAL 786', 'RAL 382', 'RAL 395',  'RAL 646', 'RAL 776', 'RAL 129', 'RAL 370', 'RAL 799', 'RAL 486', 'RAL 737',  'RAL 443')

Data_updated$Treatment<- factor(Data_updated$Treatment, levels = c('NI', 'I'))

ggplot(Data_updated, aes(x=factor(Genotype, level=level_order), y = Total_eggs, color= Treatment)) + geom_boxplot() +labs (x= "", y= "Average number of eggs per female") + theme_bw() + theme(axis.text.x = element_text(angle = 90))
```



#### Test for difference in egg production between treatments within genotypes
```{r}
model <- lmer(Total_eggs ~ Genotype * Treatment + (1|Block), Data_updated)
```


#### Calculating Estimated marginal (EM) means for Treatment within Genotype
```{r}
emm <- emmeans(model, ~ Treatment | Genotype)
```


#### Pairwise comparisons of treatments within each genotype
```{r}
pairs(emm)
```



#### Generate a dataframe containing a single EM means value per genotype for correlation study
```{r}
emm1 <- emmeans(model, ~ Genotype * Treatment)
summary_df <- as.data.frame(emm1)
```



#### Bringing in the survival data for correlation
```{r}
Cor<- data.frame(Genotype= c('RAL 837', 'RAL 397', 'RAL 786', 'RAL 382', 'RAL 395',  'RAL 646', 'RAL 776', 'RAL 129', 'RAL 370', 'RAL 799', 'RAL 486', 'RAL 737',  'RAL 443'), Ovariole_number= c(12.05, 13, 13.2, 13.65, 13.9, 14.3, 14.8, 20.05, 22.95, 24.1,24.1, 24.75, 27.4), Survival_proportion= c(0.15, 0.7717391, 0.4583333, 0.4166667, 0.2947368, 0.7752809, 0.6526316, 0.3924051, 0.5280899, 0.6304348, 0.3222222,   0.3974359,0.3974359))
Cor$Genotype<- as.factor(Cor$Genotype)
```



#### Merge two datsets and subset for each treatment condition
```{r}
Modified <- merge(Cor, summary_df, by = "Genotype")
Modified_Infected <- subset(Modified, Treatment == "I")
Modified_Uninfected <- subset(Modified, Treatment == "NI")
```



#### Plot correlation between ovariole number and egg production in infected females
```{r}
ggscatter(Modified_Infected,  x = "Ovariole_number", y = "emmean",
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman") + xlab("Ovariole number") + ylab("Number of eggs")
```



#### Correlation between ovariole number and egg production in uninfected females
```{r}
ggscatter(Modified_Uninfected,  x = "Ovariole_number", y = "emmean",
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman") + xlab("Ovariole number") + ylab("Number of eggs")
```



#### Correlation between ovariole number and survival in infected females
```{r}
ggscatter(Modified_Infected,  x = "Ovariole_number", y = "Survival_proportion",
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman") + xlab("Ovariole number") + ylab("Survival proportion")
```



#### Correlation between egg number and survival in infected females
```{r}
ggscatter(Modified_Infected,  x = "emmean", y = "Survival_proportion",
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman") + xlab("Number of eggs") + ylab("Survival proportion")
```


#### Per day egg production plot
```{r}
Data_per_day<- Data %>% group_by(Genotype, Treatment, Day, Tube, Block) %>% 
        mutate(Total = sum(Eggs_per_female)) %>% group_by(Genotype, Treatment, Day, Block) %>% mutate(Mean = mean(Total)) %>% distinct(Genotype, Treatment,Total)

level_order<- c('RAL 837', 'RAL 397', 'RAL 786', 'RAL 382', 'RAL 395',  'RAL 646', 'RAL 776', 'RAL 129', 'RAL 370', 'RAL 799', 'RAL 486', 'RAL 737',  'RAL 443')
Data_per_day$Treatment<- factor(Data_per_day$Treatment, levels = c('NI', 'I'))

ggplot(Data_per_day, aes(x=factor(Genotype, level=level_order), y = Total, color= Treatment)) + geom_boxplot() +labs (x= "", y= "Average number of eggs per female") + theme_bw() +facet_grid(rows = vars(Day)) + theme(axis.text.x = element_text(angle = 90))
```





### Offspring data

#### Plot offspring number for all genotypes
```{r}
ggplot(Data_updated, aes(x=factor(Genotype, level=level_order), y = Total_offspring, color= Treatment)) + geom_boxplot() +labs (x= "", y= "Average number of offspring per female") + theme_bw() + theme(axis.text.x = element_text(angle = 90))
```


#### Test whether there is difference in offspring number between treatments within genotypes
```{r}
model_offspring <- lmer(Total_offspring ~ Genotype * Treatment + (1|Block), Data_updated)
```


#### Calculating EM means for Treatment within Genotype
```{r}
emm_offspring <- emmeans(model_offspring, ~ Treatment | Genotype)
```


#### Pairwise comparisons of treatments within each genotype
```{r}
pairs(emm_offspring)
```


#### In order to get single EMl means value for each genotype for correlation study
```{r}
emm_offspring1 <- emmeans(model_offspring, ~ Genotype * Treatment)
offspring_summary_df <- as.data.frame(emm_offspring1)
```


#### Merging the data with ovariole number data from above
```{r}
Offspring_Modified <- merge(Cor, offspring_summary_df, by = "Genotype")
Offspring_Modified_Infected <- subset(Offspring_Modified, Treatment == "I")
Offspring_Modified_Uninfected <- subset(Offspring_Modified, Treatment == "NI")
```


#### Plot correlation between ovariole number and offspring count in infected females
```{r}
ggscatter(Offspring_Modified_Infected,  x = "Ovariole_number", y = "emmean",
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman") + xlab("Ovariole number") + ylab("Number of offspring")
```


#### Plot correlation between ovariole number and offspring count in uninfected females
```{r}
ggscatter(Offspring_Modified_Uninfected,  x = "Ovariole_number", y = "emmean",
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman") + xlab("Number of eggs") + ylab("Survival proportion")
```


```{r}
sessionInfo()
```
